---
title: "My Example Document"
author: "Brandon Kopp"
date: "7/13/2019"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Fisher's Iris Data
#### by Brandon Kopp

The famous **Fisher iris data set** gives the measurements in centimeters of the variables sepal length and width and petal length and width, respectively, for 50 flowers from each of 3 species of iris. The species are Iris *setosa*, *versicolor*, and *virginica*.

`iris` is a data frame with `r nrow(iris)` cases (rows) and `r ncol(iris)` variables (columns) named: 

1. Sepal.Length
2. Sepal.Width
3. Petal.Length
4. Petal.Width
5. Species

For more information about this data set, see [Fisher's Iris Data Set on Wikipedia](https://en.wikipedia.org/wiki/Iris_flower_data_set).

### Load Libraries

Load the `dplyr` and `ggplot2` packages for later use.

```{r warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
```

### Data Transformation

Calculate new variables that represent the area of the Sepals and Petals (Length X Width).

```{r}
iris_df <- iris %>%
  mutate(sepal_area = Sepal.Length * Sepal.Width,
         petal_area = Petal.Length * Petal.Width)
```


### Data Table

Display the data using a table.

```{r}
knitr::kable(head(iris_df))
```

### Plot

Create a scatterplot of the Sepal Area and Petal Area variables calculated above.

```{r fig.align='center', fig.width=6, fig.height=4, dpi=150}
ggplot(iris_df, aes(sepal_area, petal_area)) +
  geom_point(aes(color=Species), alpha=0.6, size=3) +
  labs(x="Sepal Area (in sq cm)", y="Petal Area (in sq cm)") +
  theme_bw()
```

